---
title: "Lab 9 - HPC"
output: 
  # pdf_document: default
  # html_document: default
  tufte::tufte_html:
    css: style.css
link-citations: yes
---

# Learning goals

In this lab, you are expected to learn/put in practice the following skills:

- Evaluate whether a problem can be parallelized or not.
- Practice with the parallel package.
- Use Rscript to submit jobs.

```{r eval=FALSE, echo=FALSE}
# install any missing packages
install.packages("microbenchmark")
```

## Problem 1: Think

Give yourself a few minutes to think about what you learned about parallelization. List three
examples of problems that you believe may be solved using parallel computing,
and check for packages on the HPC CRAN task view that may be related to it.

_Answer here._ 1. Image Processing: Processing large image datasets for tasks like feature extraction, object recognition, or image enhancement can be computationally intensive. We can break the image into sections and process each section concurrently using different processors.
2. Summing a Large List of Numbers: Adding up a huge list of numbers can be time-consuming if done sequentially. We can divide the list into smaller chunks and have multiple processors or cores summing up each chunk simultaneously.
3. Searching in a Database: Searching a massive database for specific entries can take a long time in a single-threaded approach. We can divide the database into partitions and search each partition concurrently to speed up the overall search process.

Related packages: 

## Problem 2: Pre-parallelization

The following functions can be written to be more efficient without using
`parallel`:

1. This function generates a `n x k` dataset with all its entries having a Poisson distribution with mean `lambda`.

```{r p2-fun1, eval = TRUE}
fun1 <- function(n = 100, k = 4, lambda = 4) {
  x <- NULL
  
  for (i in 1:n)
    x <- rbind(x, rpois(k, lambda))
  
  return(x)
}

fun1alt <- function(n = 100, k = 4, lambda = 4) {
  matrix(rpois(n*k, lambda=lambda), ncol=k)
}

# Benchmarking
microbenchmark::microbenchmark(
  fun1(100),
  fun1alt(100)
)
```

How much faster?

_Answer here._ fun1alt is about ten times faster than fun1 according to their mean runtime.


2.  Find the column max (hint: Checkout the function `max.col()`).

```{r p2-fun2, eval = TRUE}
# Data Generating Process (10 x 10,000 matrix)
set.seed(1234)
x <- matrix(rnorm(1e4), nrow=10)

# Find each column's max value
fun2 <- function(x) {
  apply(x, 2, max)
}

fun2alt <- function(x) {
  x[cbind(max.col(t(x)),1:col(x))]
}

# Benchmarking
bench<-microbenchmark::microbenchmark(
  fun2(x),
  fun2alt(x)
)
```

_Answer here with a plot._
```{r, eval = TRUE}
boxplot(bench, 
        main = "Box Plot for Functions",
        xlab = "Function",
        ylab = "Time",
        col = c("skyblue", "lightcoral"),
        names = c("fun2", "fun2alt"))
```


## Problem 3: Parallelize everything

We will now turn our attention to non-parametric 
[bootstrapping](https://en.wikipedia.org/wiki/Bootstrapping_(statistics)).
Among its many uses, non-parametric bootstrapping allow us to obtain confidence
intervals for parameter estimates without relying on parametric assumptions.

The main assumption is that we can approximate many experiments by resampling
observations from our original dataset, which reflects the population. 

This function implements the non-parametric bootstrap:

```{r p3-boot-fun, eval = TRUE}
library(parallel)
my_boot <- function(dat, stat, R, ncpus = 1L) {
  
  # Getting the random indices
  n <- nrow(dat)
  idx <- matrix(sample.int(n, n*R, TRUE), nrow=n, ncol=R)
 
  # Making the cluster using `ncpus`
  # STEP 1: GOES HERE
  cl <- makePSOCKcluster(ncpus)
  # STEP 2: GOES HERE
  clusterExport(cl, varlist=c("idx", "dat", "stat"), envir = environment())
  
  # STEP 3: THIS FUNCTION NEEDS TO BE REPLACED WITH parLapply
  # ans <- lapply(seq_len(R), function(i) {
  #   stat(dat[idx[,i], , drop=FALSE])
  # })
  
  ans <- parLapply(cl, seq_len(R), function(i) {
    stat(dat[idx[,i], , drop=FALSE])
  })
  
  # Coercing the list into a matrix
  ans <- do.call(rbind, ans)
  
  # STEP 4: GOES HERE
  stopCluster(cl)
  ans
  
}
```

1. Use the previous code, and make it work with `parallel`. Here is just an example
for you to try:

```{r p3-test-boot, eval = TRUE}
# Bootstrap of a linear regression model
my_stat <- function(d) coef(lm(y~x,data=d))

# DATA SIM
set.seed(1)
n <- 500 
R <- 1e4
x <- cbind(rnorm(n)) 
y <- x*5 + rnorm(n)

# Check if we get something similar as lm
ans0 <- confint(lm(y~x))
ans1 <- my_boot(dat=data.frame(x,y), stat = my_stat, R=R, ncpus=4)

ans0
t(apply(ans1, 2, quantile, probs=c(.025,.975)))
```

2. Check whether your version actually goes faster than the non-parallel version:

```{r benchmark-problem3, eval = TRUE}
# your code here
parallel::detectCores()
system.time(my_boot(dat=data.frame(x,y), stat = my_stat, R=R, ncpus=1))
system.time(my_boot(dat=data.frame(x,y), stat = my_stat, R=R, ncpus=8))
```

_Answer here._
Non-parallel version runs in 6 seconds whereas parallel version runs in 3 seconds, which is about 2 times faster.

## Problem 4: Compile this markdown document using Rscript

Once you have saved this Rmd file, try running the following command
in your terminal:

```bash
Rscript --vanilla -e 'rmarkdown::render("lab09-hpc.Rmd")' &
```

Where `[full-path-to-your-Rmd-file.Rmd]` should be replace with the full path to
your Rmd file... :).


