---
title: "Lab 08 - Text Mining/NLP"
output: github_document
author: "Cathy Pei"
date: "Feb 28, 2024"
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(eval = F, include  = T)
```

# Learning goals

- Use `unnest_tokens()` and `unnest_ngrams()` to extract tokens and ngrams from text
- Use dplyr and ggplot2 to analyze and visualize text data
- Try a theme model using `topicmodels`

# Lab description

For this lab we will be working with the medical record transcriptions from https://www.mtsamples.com/. And is loaded and "fairly" cleaned at https://github.com/JSC370/JSC370-2024/tree/main/data/medical_transcriptions.

This markdown document should be rendered using `github_document` document.



### Setup packages

You should load in `dplyr`, (or `data.table` if you want to work that way), `ggplot2` and `tidytext`. Install `wordcloud`, `tm`, and `topicmodels` if you don't alreadyh have them.



### Read in the Medical Transcriptions

Loading in reference transcription samples from https://www.mtsamples.com/

```{r eval = TRUE}
library(tidytext)
library(tidyverse)
library(wordcloud)
library(tm)
library(topicmodels)
library(ggplot2)

mt_samples <- read_csv("https://raw.githubusercontent.com/JSC370/JSC370-2024/main/data/medical_transcriptions/mtsamples.csv")
mt_samples <- mt_samples %>%
  select(description, medical_specialty, transcription)

head(mt_samples)
```

---

## Question 1: What specialties do we have?

We can use `count()` from `dplyr` to figure out how many different medical specialties are in the data. Are these categories related? overlapping? evenly distributed? Make a bar plot.

```{r eval = TRUE}
mt_samples |>
  count(medical_specialty, sort = TRUE)|> 
  ggplot(aes(x = reorder(medical_specialty, n), y = n, fill = medical_specialty)) +
  geom_bar(stat = "identity") +
  labs(title = "Distribution of Medical Specialties", x = "Medical Specialty", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

#### Observation
The categories are quite diverse, with no overlap (no categories share the same medical specialty). They are not evenly distributed, as we can see that surgery appears to have a significant number of counts compared to other medical specialties.

---

## Question 2: Tokenize

- Tokenize the the words in the `transcription` column
- Count the number of times each token appears
- Visualize the top 20 most frequent words with a bar plot
- Create a word cloud

Explain what we see from this result. Does it makes sense? What insights (if any) do we get?

```{r eval = TRUE}
# Tokenize the the words in the `transcription` column
tokens <- mt_samples %>%
  select(transcription) %>%
  unnest_tokens(word, transcription)

# Count the number of times each token appears
token_counts <- tokens %>%
  count(word, sort = TRUE)

# Visualize the top 20 most frequent words with a bar plot
token_counts_top20 <- head(token_counts, 20)

token_counts_top20 %>%
  ggplot(aes(x = reorder(word, n), y = n, fill = word)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 20 Most Frequent Words", x = "Words", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# Create a word cloud
wordcloud(words = token_counts$word[1:500], freq = token_counts$n[1:500], scale = c(3, 0.5), colors = brewer.pal(8, "Dark2"))
```

# Observation
The top 20 words we got, such as "the," "and," "was," etc., are common English stop words. Stop words are words that are commonly used in a language but are typically filtered out when analyzing text data because they don't carry significant meaning on their own. We should remove them to focus on more meaningful content words.

---

## Question 3: Stopwords

- Redo Question 2 but remove stopwords(use stopwords package)
- Bonus points if you remove numbers as well (use regex)

What do we see know that we have removed stop words? Does it give us a better idea of what the text is about?

- Try customizing your stopwords list to include 3-4 additional words that do not appear informative

```{r eval = TRUE}
head(stopwords("english"))
length(stopwords("english"))

# Add custom stopwords
custom_stopwords <- c("left", "right", "normal", "well", "placed", "history", "patient", "using", "used", "will")
numeric_pattern <- paste0("\\b", 0:9, "\\b")
stop_words <- unique(c(stopwords("english"), custom_stopwords, numeric_pattern))

# Tokenize the the words in the `transcription` column with the remove of stopwords
tokens <- mt_samples %>%
  select(transcription) %>%
  unnest_tokens(word, transcription) %>%
  mutate(word = as.character(word)) %>%  # Convert 'word' to character
  filter(!(word %in% stop_words) & !grepl("\\b\\d+\\b", word))

token_counts <- tokens %>%
  count(word, sort = TRUE)

# Visualize the top 20 most frequent words with a bar plot
token_counts_top20 <- head(token_counts, 20)

# Bar plot
token_counts_top20 %>%
  ggplot(aes(x = reorder(word, n), y = n, fill = word)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 20 Most Frequent Words (Stopwords Removed)", x = "Words", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Create a word cloud for the top 100 most frequent words
wordcloud(words = token_counts$word[1:100], freq = token_counts$n[1:100], scale = c(2, 0.5), colors = brewer.pal(8, "Dark2"))
```

#### Observation
By removing the stopwords, we can observe more informative words such as "blood," "diagnosis," and "anesthesia," which are more closely related to the medical context.
---



# Question 4: ngrams

Repeat question 2, but this time tokenize into bi-grams. How does the result change if you look at tri-grams? Note we need to remove stopwords a little differently. You don't need to recreate the wordclouds.

```{r eval = TRUE}
custom_stopwords <- c("left", "right", "normal", "well", "placed", "history", "patient", "using", "used", "will")
stopwords2 <- unique(c(stopwords("english"), custom_stopwords))

sw_start <- paste0("^", paste(stopwords2, collapse=" |^"), "$")
sw_end <- paste0("", paste(stopwords2, collapse="$| "), "$")


# Tokenize into bigrams and remove stopwords
tokens_bigram <- mt_samples %>%
  select(transcription) %>%
  unnest_tokens(bigram, transcription, token = "ngrams", n = 2) 
tokens_bigram <- tokens_bigram %>%
  filter(!grepl(sw_start, tokens_bigram$bigram, ignore.case = TRUE) & 
    !grepl(sw_end, tokens_bigram$bigram, ignore.case = TRUE))
  
token_counts <- tokens_bigram %>%
  count(bigram, sort = TRUE)

# Visualize the top 20 most frequent words with a bar plot
token_counts_top20 <- head(token_counts, 20)

# Bar plot
token_counts_top20 %>%
  ggplot(aes(x = reorder(bigram, n), y = n, fill = bigram)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 20 Most Frequent Bigrams (Stopwords Removed)", x = "Bigrams", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r eval = TRUE}
custom_stopwords <- c("left", "right", "normal", "well", "placed", "history", "patient", "using", "used", "will")
stopwords2 <- unique(c(stopwords("english"), custom_stopwords))

sw_start <- paste0("^", paste(stopwords2, collapse=" |^"), "$")
sw_end <- paste0("", paste(stopwords2, collapse="$| "), "$")


# Tokenize into bigrams and remove stopwords
tokens_trigram <- mt_samples %>%
  select(transcription) %>%
  unnest_tokens(trigram, transcription, token = "ngrams", n = 3) 
tokens_trigram <- tokens_trigram %>%
  filter(!grepl(sw_start, tokens_trigram$trigram, ignore.case = TRUE) & 
    !grepl(sw_end, tokens_trigram$trigram, ignore.case = TRUE))
  
token_counts <- tokens_trigram %>%
  count(trigram, sort = TRUE)

# Visualize the top 20 most frequent words with a bar plot
token_counts_top20 <- head(token_counts, 20)

# Bar plot
token_counts_top20 %>%
  ggplot(aes(x = reorder(trigram, n), y = n, fill = trigram)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 20 Most Frequent Trigrams (Stopwords Removed)", x = "Trigrams", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Observation
We can see that, compared to bigrams, trigrams are more interpretable and relatable to the medical context, as trigrams form more professional phrases.
---

# Question 5: Examining words

Using the results from the bigram, pick a word and count the words that appear before and after it, and create a plot of the top 20.

```{r eval = TRUE}
library(stringr)
# Use regex to filter bigrams containing the word "blood"
blood_bigrams <- tokens_bigram %>%
  filter(str_detect(bigram, "\\bblood\\b"))

# Tokenize the bigrams into unigrams
tokens_unigram <- blood_bigrams %>%
  unnest_tokens(word, bigram, token = "ngrams", n = 1) %>%
  filter(!(word == "blood"))  # Filter to only include words that are not "blood"

token_counts <- tokens_unigram %>%
  count(word, sort = TRUE)

# Visualize the top 20 most frequent words with a bar plot
token_counts_top20 <- head(token_counts, 20)

# Bar plot
token_counts_top20 %>%
  ggplot(aes(x = reorder(word, n), y = n, fill = word)) +
  geom_bar(stat = "identity") +
  labs(title = "Top 20 Most Frequent Words Before or After Blood (Stopwords Removed)", x = "Words", y = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

---


# Question 6: Words by Specialties

Which words are most used in each of the specialties. you can use `group_by()` and `top_n()` from `dplyr` to have the calculations be done within each specialty. Remember to remove stopwords. How about the most 5 used words?


```{r eval = TRUE}
custom_stopwords <- c("left", "right", "normal", "well", "placed", "history", "patient", "using", "used", "will")
stopwords2 <- unique(c(stopwords("english"), custom_stopwords))

# Tokenize the words in the 'transcription' column and remove stopwords
tokens <- mt_samples %>%
  select(medical_specialty, transcription) %>%
  unnest_tokens(word, transcription) %>%
  mutate(word = as.character(word)) %>%  # Convert 'word' to character
  filter(!(word %in% stop_words) & !grepl("\\b\\d+\\b", word))

# Group by specialty and find the top 5 used words in each group
top_words_per_specialty <- tokens %>%
  group_by(medical_specialty, word) %>%
  summarize(count = n()) %>%
  group_by(medical_specialty) %>%
  top_n(5, count) %>%
  arrange(medical_specialty, desc(count))

# Print the top words for each medical specialty
print(top_words_per_specialty)
```


# Question 7: Topic Models

See if there are any themes in the data by using a topic model (LDA). 

- you first need to create a document term matrix
- then you can try the LDA function in `topicmodels`. Try different k values.
- create a facet plot of the results from the LDA (see code from lecture)


```{r eval = TRUE}

# Tokenize the words in the 'transcription' column and remove stopwords
tokens <- mt_samples %>%
  select(transcription) %>%
  unnest_tokens(word, transcription) %>%
  mutate(word = as.character(word)) %>%  # Convert 'word' to character
  filter(!(word %in% stop_words) & !grepl("\\b\\d+\\b", word))

# Create a document-term matrix
transcripts_dtm <- tokens %>%
  count(document = 1, word) %>%
  cast_dtm(document = document, term = word, value = n)

# Convert DTM to a matrix
transcripts_matrix <- as.matrix(transcripts_dtm)

# Create a function to fit LDA model and extract topics
fit_lda <- function(k) {
  lda_model <- LDA(transcripts_dtm, k = k, control = list(seed = 1234))
  topics <- tidy(lda_model, matrix = "beta")
  # Filter top 20 terms for each topic
  top_terms <- topics %>%
    group_by(topic, term) %>%
    top_n(1, wt = beta) %>%
    ungroup() %>%
    top_n(20, wt = desc(beta))
  return(top_terms)
}

# Try different k values
k_values <- c(3, 5, 7)

# Fit LDA models for different k values
lda_results <- lapply(k_values, fit_lda)

# Create a facet plot of the results
facet_plot <- lda_results %>%
  bind_rows(.id = "k") %>%
  ggplot(aes(x = reorder(term, beta), y = beta, fill = factor(k))) +
  geom_bar(stat = "identity") +
  facet_wrap(~k, scales = "free_y", ncol = 1) +
  labs(title = "Top 20 Terms for Each Topic with Different k Values", x = "Terms", y = "Beta") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

print(facet_plot)

```



# Deliverables

1. Questions 1-7 answered, raw .Rmd file and pdf or html output uploaded to Quercus
